---
title: "Homework 3"
author: "Pin Wang"
date: "2025-05-27"
weight: 3
callout-appearance: minimal
execute:
  python: ./.venv/bin/python  
  echo: true
  warning: false
  message: false
format:
  html:
    code-fold: true
    theme: cerulean
---

This assignment explores two methods for estimating the Multinomial Logit (MNL) model:  
1. **Maximum Likelihood Estimation (MLE)**  
2. **Bayesian Estimation using Metropolis-Hastings MCMC**  

We use simulated conjoint data from a hypothetical survey on video streaming service preferences.

---

## 1. Likelihood for the MNL Model

We consider $n$ consumers each selecting one product $j$ from $J$ alternatives. The utility they derive from alternative $j$ is:

$$ U_{ij} = x_j'\beta + \varepsilon_{ij} $$

Assuming $\varepsilon_{ij}$ follows an i.i.d. Gumbel distribution leads to the standard MNL formulation:

$$ \mathbb{P}_i(j) = \frac{\exp(x_j'\beta)}{\sum_{k=1}^J \exp(x_k'\beta)} $$

The log-likelihood across all consumers is:

$$ \ell(\beta) = \sum_{i=1}^n \sum_{j=1}^J \delta_{ij} \log(\mathbb{P}_i(j)) $$

---

## 2. Simulated Conjoint Data

We simulate a dataset of 100 consumers, each completing 10 choice tasks among 3 alternatives. Each alternative has:

- Brand: Netflix (N), Prime (P), or Hulu (H)
- Ads: Yes or No
- Price: \$8–\$32

Part-worth utility parameters:

- $\beta_{Netflix}=1.0$, $\beta_{Prime}=0.5$ (Hulu = baseline)
- $\beta_{Ads} = -0.8$
- $\beta_{Price} = -0.1$

```{r}
set.seed(123)
brand <- c("N", "P", "H")
ad <- c("Yes", "No")
price <- seq(8, 32, by=4)

profiles <- expand.grid(brand = brand, ad = ad, price = price)
m <- nrow(profiles)

b_util <- c(N = 1.0, P = 0.5, H = 0)
a_util <- c(Yes = -0.8, No = 0.0)
p_util <- function(p) -0.1 * p

n_peeps <- 100
n_tasks <- 10
n_alts <- 3

sim_one <- function(id) {
    datlist <- list()
    for (t in 1:n_tasks) {
        dat <- cbind(resp = id, task = t, profiles[sample(m, size = n_alts), ])
        dat$v <- b_util[dat$brand] + a_util[dat$ad] + p_util(dat$price)
        dat$e <- -log(-log(runif(n_alts)))
        dat$u <- dat$v + dat$e
        dat$choice <- as.integer(dat$u == max(dat$u))
        datlist[[t]] <- dat
    }
    do.call(rbind, datlist)
}

conjoint_data <- do.call(rbind, lapply(1:n_peeps, sim_one))
```

---

## 3. Preparing the Data for Estimation

We encode categorical variables into binary features and prepare the design matrix for estimation.

```{r}
conjoint_data$brand_N <- ifelse(conjoint_data$brand == "N", 1, 0)
conjoint_data$brand_P <- ifelse(conjoint_data$brand == "P", 1, 0)
conjoint_data$ads <- ifelse(conjoint_data$ad == "Yes", 1, 0)
```

---

## 4. Estimation via Maximum Likelihood

We estimate the model by maximizing the log-likelihood function using `optim()`:

```{r}
log_lik <- function(beta, data) {
  b_N <- beta[1]
  b_P <- beta[2]
  b_ads <- beta[3]
  b_price <- beta[4]

  data$utility <- with(data, b_N * brand_N + b_P * brand_P + b_ads * ads + b_price * price)
  data$exp_util <- exp(data$utility)
  data$sum_exp_util <- ave(data$exp_util, data$resp, data$task, FUN = sum)
  data$prob <- data$exp_util / data$sum_exp_util

  ll <- sum(log(data$prob[data$choice == 1]))
  return(-ll)
}

start_vals <- c(0, 0, 0, 0)
mle_fit <- optim(start_vals, log_lik, data = conjoint_data, hessian = TRUE)
```

MLE results and confidence intervals:

```{r}
estimates <- mle_fit$par
se <- sqrt(diag(solve(mle_fit$hessian)))
ci <- cbind(estimates - 1.96 * se, estimates + 1.96 * se)
mle_results <- data.frame(Parameter = c("Netflix", "Prime", "Ads", "Price"),
                          Estimate = estimates, SE = se,
                          CI_Lower = ci[,1], CI_Upper = ci[,2])
knitr::kable(mle_results, digits = 3)
```

---

## 5. Bayesian Estimation via MCMC

We implement Metropolis-Hastings sampling using normal priors:

```{r}
log_prior <- function(beta) {
  sum(dnorm(beta[1:3], 0, 5, log = TRUE)) + dnorm(beta[4], 0, 1, log = TRUE)
}

log_posterior <- function(beta, data) {
  -log_lik(beta, data) + log_prior(beta)
}

run_mcmc <- function(start, steps, data) {
  current <- start
  samples <- matrix(NA, nrow = steps, ncol = 4)
  for (s in 1:steps) {
    proposal <- current + c(rnorm(3, 0, 0.05), rnorm(1, 0, 0.005))
    log_alpha <- log_posterior(proposal, data) - log_posterior(current, data)
    if (log(runif(1)) < log_alpha) current <- proposal
    samples[s, ] <- current
  }
  return(samples)
}

samples <- run_mcmc(estimates, 11000, conjoint_data)
posterior <- samples[1001:11000, ]
```

Posterior summaries and diagnostic plots:

```{r}
posterior_means <- colMeans(posterior)
posterior_sd <- apply(posterior, 2, sd)
posterior_ci <- t(apply(posterior, 2, function(x) quantile(x, c(0.025, 0.975))))

posterior_summary <- data.frame(Parameter = c("Netflix", "Prime", "Ads", "Price"),
                                Mean = posterior_means,
                                SD = posterior_sd,
                                CI_Lower = posterior_ci[,1],
                                CI_Upper = posterior_ci[,2])
knitr::kable(posterior_summary, digits = 3)

```{r}
# Trace plots for all parameters
par(mfrow=c(2,2))
plot(posterior[,1], type="l", main="Trace: Beta Netflix", ylab="Value", xlab="Iteration")
plot(posterior[,2], type="l", main="Trace: Beta Prime", ylab="Value", xlab="Iteration")
plot(posterior[,3], type="l", main="Trace: Beta Ads", ylab="Value", xlab="Iteration")
plot(posterior[,4], type="l", main="Trace: Beta Price", ylab="Value", xlab="Iteration")

# Posterior histograms for all parameters
par(mfrow=c(2,2))
hist(posterior[,1], breaks=30, main="Posterior: Beta Netflix", xlab="Value")
hist(posterior[,2], breaks=30, main="Posterior: Beta Prime", xlab="Value")
hist(posterior[,3], breaks=30, main="Posterior: Beta Ads", xlab="Value")
hist(posterior[,4], breaks=30, main="Posterior: Beta Price", xlab="Value")
```
```

---

## 6. Discussion

- **Interpretation**:  
  $\beta_{Netflix} > \beta_{Prime}$ implies respondents prefer Netflix over Prime.  
  The negative $\beta_{Price}$ indicates consumers value lower prices.

- **MLE vs. Bayesian**:  
  Results are consistent across methods. MCMC posterior means and intervals closely match MLE estimates, increasing confidence in inference.

- **Extension**:  
  To handle real-world heterogeneity, we can extend the MNL to a **Hierarchical Bayes Model**, introducing individual-level betas drawn from a population distribution (e.g., $\beta_i \sim \mathcal{N}(\mu, \Sigma)$), allowing personalization and richer insights.
